---
title: "Black Friday Data Visualization"
author: "Parv"
date: "8 March 2018"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Problem Statement

To understand the customer purchase behaviour (specifically, purchase amount) against various products of different categories so that sales of a retail company can be boosted.

We will be building models to predict the purchase amount of customer against various products which will help us to create personalized offer for customers against different products.

## 2. Setting up the environment

Setting up environment for analysis, loading data, packages, understanding variables.

## 2.1 Loading Data

```{r}
getwd()
data_set<-read.csv(paste("train.csv",sep = ""))

## 75% of the sample size
smp_size <- floor(0.75 * nrow(data_set))

## setting the seed to make partition reproductible
set.seed(123)
train_ind <- sample(seq_len(nrow(data_set)), size = smp_size)

train_set <- data_set[train_ind, ]
test_set <- data_set[-train_ind, ]

y_test<-as.data.frame(test_set[,12], drop=false)
names(y_test)<-c("purchase")

test_set<-as.data.frame(test_set[,1:11], drop=false)
dim(test_set)
```

## 2.2 Checking dimensions of the data

```{r}
dim(train_set)
dim(test_set)
```

**Training data** set consists of 4,12,551 entries across 12 variables.  
**Test Data** set consists of 1,37,517 entries across 11 variables.

12th variable in Training Data set is our response Variable which needs to be predicted in test data set.

# 2.3 Understanding the Data set

```{r}
library(psych)
describe(train_set)
describe(test_set)
```

# 3. Univariate Analysis

Let's First Find categorical and continuos variables in our training data set.

```{r}
str(train_set)
```

## Data Prepartion & Exploratory Data Analysis 
Firstly doing the univariate exploration and modifying the data if deemed necessary. 

**one way contigency table**

```{r}
mytable <- with(train_set,table(Gender))
mytable
lbls <- c("M","F")
pct <- round(mytable/sum(mytable)*100)
lbls <- paste(lbls, pct) 
lbls <- paste(lbls,"%",sep="")
pie(mytable,labels = lbls)
```

```{r}
mytable1 <- with(train_set,table(Marital_Status))
mytable1
lbls1 <- c("Single","Married")
pct1 <- round(mytable1/sum(mytable1)*100)
lbls1 <- paste(lbls1, pct1) 
lbls1<- paste(lbls1,"%",sep="")
pie(mytable1,labels = lbls1)
```

```{r}
mytable2 <- with(train_set,table(City_Category))
mytable2
lbls2 <- c("A","B","C")
pct2 <- round(mytable2/sum(mytable2)*100)
lbls2 <- paste(lbls2, pct2) 
lbls2<- paste(lbls2,"%",sep="")
pie(mytable2,labels = lbls2)
```


```{r}
par(mfrow = c(1, 2))
#train data
with(train_set,table(Stay_In_Current_City_Years))
#test data
with(test_set,table(Stay_In_Current_City_Years))
```

```{r}
par(mfrow = c(1, 2))
a<-with(train_set,table(Product_Category_1))
#a
barplot(a,main ="train_data")
b<-with(test_set,table(Product_Category_1))
#b
barplot(b,main ="test_data")
```

```{r}
par(mfrow = c(1, 2))
a<-with(train_set,table(Product_Category_2))
#a
barplot(a,main ="train_data")
b<-with(test_set,table(Product_Category_2))
#b
barplot(b,main ="test_data")
```

```{r}
library(lattice)
par(mfrow = c(1, 2))
a<-with(train_set,table(Product_Category_3))
#a
barchart(a,main ="train_data",horizontal = "FALSE")
b<-with(test_set,table(Product_Category_3))
#b
barchart(b,main ="test_data",horizontal = "FALSE")
```

**2 way table**

```{r}
xtabs(~Marital_Status+Gender,data=train_set)
xtabs(Purchase~Product_Category_3,aggregate(Purchase~Product_Category_3,train_set,mean))
xtabs(Purchase~Product_Category_2,aggregate(Purchase~Product_Category_2,train_set,mean))
xtabs(Purchase~Product_Category_1,aggregate(Purchase~Product_Category_1,train_set,mean))
```

## 3.1. Unique Data For EDA & Data type modification 
Since our data is stored based on product id i.e. if a person buying 10 products then his data will be stored in 10 observation and hence there will cause a repetition of same person data.

```{r}
train_set$User_ID <- as.factor(train_set$User_ID)
train_set$Product_ID <- as.factor(train_set$Product_ID)
train_set$Marital_Status <- as.factor(ifelse(train_set$Marital_Status == 1, 'Married', 'Single'))
train_set$Age <- as.factor(train_set$Age)
train_set$Gender <- as.factor(ifelse(train_set$Gender=='M', 'Male', 'Female'))
train_set$Occupation <- as.factor(train_set$Occupation)
train_set$City_Category <- as.factor(train_set$City_Category)
train_set$Stay_In_Current_City_Years <- as.factor(train_set$Stay_In_Current_City_Years)

test_set$User_ID <- as.factor(test_set$User_ID)
test_set$Product_ID <- as.factor(test_set$Product_ID)
test_set$Marital_Status <- as.factor(ifelse(test_set$Marital_Status == 1, 'Married', 'Single'))
test_set$Age <- as.factor(test_set$Age)
test_set$Gender <- as.factor(ifelse(test_set$Gender=='M', 'Male', 'Female'))
test_set$Occupation <- as.factor(test_set$Occupation)
test_set$City_Category <- as.factor(test_set$City_Category)
test_set$Stay_In_Current_City_Years <- as.factor(test_set$Stay_In_Current_City_Years)

#str(train_set)
#str(test_set)
```

The function distinct() in dplyr package can be used to keep only unique/distinct rows from a data frame. If there are duplicate rows, only the first row is preserved. It’s an efficient version of the R base function unique(). 

```{r}
library("dplyr")
EDA_Distinct <- distinct(train_set, User_ID, Age, Gender, Marital_Status, Occupation, City_Category, Stay_In_Current_City_Years)
#str(EDA_Distinct)
head(EDA_Distinct)
```

#3.2. User_ID 

**Creating new variables User_Purchase_Count, Total_Spending and Avg_Spending_on_each_Product**

**Total number of product purchased bu customer**
```{r}
#creating a new data frame to stor the number of purchase made by each user
userIDCount <- as.data.frame(table(train_set$User_ID))
names(userIDCount) <- c("User_ID","User_Purchase_Count")
head(userIDCount)
```

merging df
```{r}
train_set <- merge(x = train_set, y = userIDCount, by = "User_ID", all.x = TRUE)
```

changing the test dataset too
```{r}
# writing code such that if a new user comes for the first time his count is set to one in test dataset
test_set <- merge(x = test_set, y = userIDCount, by = "User_ID", all.x = TRUE)
test_set[is.na(test_set$User_Purchase_Count), "User_Purchase_Count"] <- 1
#Changing the datatype
test_set$User_Purchase_Count <- as.integer(test_set$User_Purchase_Count)
```

**Total Spending by a User**

Changing the Training Dataset
```{r}
# creating a data frame to store the total spending by a user
totspend <- aggregate(train_set$Purchase, by=list(Category=train_set$User_ID), FUN=sum)
names(totspend) <- c("User_ID","Total_Spending")
train_set <- merge(x = train_set, y = totspend, by = "User_ID", all.x = TRUE)
#head(train_set)
```

changing test too
```{r}
test_set <- merge(x = test_set, y = totspend, by = "User_ID", all.x = TRUE)
# writing code such that if a new user comes for the first time his count is set to zero for total spending in test dataset
test_set[is.na(test_set$Total_Spending), "Total_Spending"] <- 0
#Changing the datatype
test_set$Total_Spending <- as.integer(test_set$Total_Spending)
```

**Average Spending on each product by a customer**

Now calculating the average spending of user on each product we can divide Total_Spending by User_Purchase Count.
Avg_Spending_on_each_Product = Total_Spending / User_Purchase_Count

```{r}
train_set$Avg_Spending_on_each_Product <- train_set$Total_Spending / train_set$User_Purchase_Count
test_set$Avg_Spending_on_each_Product <- test_set$Total_Spending / test_set$User_Purchase_Count
#Changing the datatype 
train_set$Avg_Spending_on_each_Product <- as.integer(train_set$Avg_Spending_on_each_Product)
test_set$Avg_Spending_on_each_Product <- as.integer(test_set$Avg_Spending_on_each_Product)
```

```{r}
summary(totspend$Total_Spending)
```

```{r}
library(ggplot2)
ggplot(totspend, aes(x=Total_Spending)) + geom_density(fill="red", col="black", alpha=0.80)
```

**removing unused df**

```{r}
rm(userIDCount, totspend)
```

# Analysing using distict data

```{r}
#Updating EDA_Distinct dataframe
EDA_Distinct <- distinct(train_set, User_ID, Age, Gender, Marital_Status, Occupation, City_Category, Stay_In_Current_City_Years, User_Purchase_Count)
d1 <- summary(EDA_Distinct$User_Purchase_Count)
d1
```

```{r}
ggplot(EDA_Distinct, aes(x=User_Purchase_Count)) + geom_density(fill="red", col="black", alpha=0.80) + annotate(geom = "text", x = 6, y = 0.0125, label = "Min")
histogram(EDA_Distinct$User_Purchase_Count)
bwplot(Age~User_Purchase_Count|Marital_Status+Gender,data = EDA_Distinct)
bwplot(User_Purchase_Count~Occupation|Marital_Status+Gender,data = EDA_Distinct)
```

#3.3 Product_id

```{r}
head(train_set$Product_ID,15)
```
i.e. there are a total of 3631 number of product

**Creating new variables Total number of product unit sold, Average price of a product, Sd of the product price**

```{r}
SoldProdCount <- as.data.frame(table(train_set$Product_ID))
names(SoldProdCount) <- c("Product_ID","Product_Sold_Count")

SoldProdPriceMean <- aggregate(train_set$Purchase, by=list(Category=train_set$Product_ID), FUN=mean)
names(SoldProdPriceMean) <- c("Product_ID","Product_Mean_Price")

SoldProdPriceSD <- aggregate(train_set$Purchase, by=list(Category=train_set$Product_ID), FUN=sd)
names(SoldProdPriceSD) <- c("Product_ID","Product_SD_Price")
```

```{r}
summary(SoldProdCount$Product_Sold_Count)
histogram(SoldProdCount$Product_Sold_Count)
histogram(SoldProdPriceMean$Product_Mean_Price)
histogram(SoldProdPriceSD$Product_SD_Price)
```

#Correlation Matrix

changing variables type to numeric

```{r}
corr_train <- train_set
corr_train$User_ID <- as.numeric(corr_train$User_ID)
corr_train$User_Purchase_Count <- as.numeric(corr_train$User_Purchase_Count)
corr_train$Product_ID <- as.numeric(corr_train$Product_ID)

corr_train$Gender <- as.numeric(ifelse(corr_train$Gender=="Male", 1, 0))

corr_train$Age <- as.numeric(ifelse(corr_train$Age=='0-17', 17, ifelse(corr_train$Age=='18-25', 25, ifelse(corr_train$Age=='26-35', 35, ifelse(corr_train$Age=='36-45', 45, ifelse(corr_train$Age=='46-50', 50, ifelse(corr_train$Age=='51-55', 55, 65)))))))

corr_train$Marital_Status <- as.numeric(ifelse(corr_train$Marital_Status=="Married", '1', '0'))

corr_train$Occupation <- as.numeric(corr_train$Occupation)

corr_train$City_Category <- as.numeric(ifelse(corr_train$City_Category=='A', 1, ifelse(corr_train$City_Category=='B', 2, 3)))

corr_train$Stay_In_Current_City_Years <- as.numeric(ifelse(corr_train$Stay_In_Current_City_Years=='4+', 6, corr_train$Stay_In_Current_City_Years))

corr_train$Total_Spending <- as.numeric(corr_train$Total_Spending)
corr_train$Avg_Spending_on_each_Product <- as.numeric(corr_train$Avg_Spending_on_each_Product)


#target Variable 
Y <- as.numeric(corr_train$Purchase)
corr_train$Product_Category_2<-NULL
corr_train$Product_Category_3<-NULL
cor(corr_train)
```

#Corrplot

```{r}
library(corrplot)
M <- cor(corr_train)
corrplot(M, method="number")
```
